\documentclass[11pt, a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{xcolor}

\geometry{margin=1in}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

\title{\Large Toward a Layered Ontology of Semantic Duality:\\
14 Candidate Dualities Across 6 Algebraic Layers\\
with Preliminary Domain and Neural Evidence}

\author{
    J. Arturo Ornelas Brand \\
    \textit{Independent Researcher} \\
    \texttt{arturoornelas62@gmail.com}
}

\date{March 2026}

\begin{document}

\maketitle

\begin{abstract}
We present a formal ontological framework in which conceptual complexity emerges from 14+ candidate dualities organized across 6 algebraic layers (Boolean, Fuzzy, Ordinal, Modal, Trivalent, Probabilistic). The framework is grounded in 72 semantic primitives connected by a directed acyclic graph (DAG) of 133+ dependency edges. We evaluate the framework through three independent methodologies: (1)~domain analysis across 8 disciplines and 3 negative controls (Astrology, Alchemy, Phrenology) using a composite 6-metric Domain Validity Score (DVS), achieving DVS~$\geq 0.719$ for all 8 positive domains while all 3 negative controls fall below the 0.50 threshold (gap = 0.233, robust across all weight configurations); (2)~Bayesian model comparison yielding a Bayes factor of 121.4 (decisive) over the reversed ordering, though only anecdotal evidence over alternative philosophical orderings (BF~=~1.14 vs.\ Hegel, 1.95 vs.\ Peirce); and (3)~neurosymbolic learning via a GPT-2 Medium model augmented with a 72-bit triadic head, which learns the primitive structure at 90.8\% bit accuracy (though this barely exceeds the 90.3\% majority-class baseline due to target sparsity) while preserving general language ability (PPL~=~31.95, identical to baseline). The learned representations exhibit near-perfect relational structure (Regla de Tres cosine~=~0.996) and show a significant phase transition during training ($p$~=~0.0004). We report both confirmatory and null results transparently: layer ordering in neural representations is not significant ($p$~=~0.71), and only 5 of 13 dual axes show significant learned anti-correlation after FDR correction.
\end{abstract}

\noindent\textbf{Keywords:} ontology, semantic primitives, emergence, duality, neurosymbolic AI, algebraic structure, domain validation

% ============================================================
\section{Introduction}

The search for universal semantic primitives---irreducible conceptual atoms from which all meaning is composed---spans philosophy (Leibniz's \emph{characteristica universalis}), linguistics (Wierzbicka's Natural Semantic Metalanguage), and computational semantics (ConceptNet, FrameNet). Existing approaches either lack formal algebraic structure or fail to demonstrate empirical universality across domains.

We propose that conceptual complexity emerges through a specific sequence of \textbf{dualities}---paired oppositions that each require all previous dualities as prerequisites. This emergence follows an algebraic chain:

$$\text{Boolean} \to \text{Fuzzy} \to \text{Ordinal} \to \text{Modal} \to \text{Trivalent} \to \text{Probabilistic}$$

Each algebra introduces operators that the previous one cannot express: degree from existence, comparison from degree, possibility from order, valuation from possibility, observation from valuation.

The contribution of this work is threefold:

\begin{enumerate}[nosep, leftmargin=*]
    \item \textbf{Formalization}: 72 semantic primitives organized in a DAG across 6 algebraic layers, with 14 dual axes and formal dependency rules.
    \item \textbf{Empirical evaluation}: cross-domain analysis (8 disciplines + 3 negative controls including Astrology, Alchemy, and Phrenology) using a composite 6-metric IDVS with statistical hardening (permutation tests, FDR correction, Bayesian model comparison, null baselines).
    \item \textbf{Neural validation}: a neurosymbolic model (GPT-2 Medium + triadic head) that learns the proposed structure from natural language, demonstrating that the algebraic relationships are recoverable from distributional semantics.
\end{enumerate}

% ============================================================
\section{The Framework}

\subsection{Semantic Primitives}

We define 72 semantic primitives distributed across 6 layers:

\begin{table}[ht]
\centering
\small
\begin{tabular}{llrl}
\toprule
\textbf{Layer} & \textbf{Algebra} & \textbf{Prims} & \textbf{Examples} \\
\midrule
1. Point (0D)    & Boolean $\{0,1\}$     & 3  & vac\'{i}o, informaci\'{o}n, uno \\
2. Line (1D)     & Fuzzy $\{0,+\}$       & 8  & fuerza, contenci\'{o}n, uni\'{o}n, separaci\'{o}n \\
3. Time (1D+t)   & Ordinal $\{<,=,>\}$   & 13 & mover, orden, caos, creaci\'{o}n \\
4. Plane (2D)    & Modal $\{\Diamond,\Box\}$ & 21 & bien, mal, verdad, mentira \\
5. Volume (3D)   & Trivalent $\{-,0,+\}$ & 23 & vida, muerte, placer, dolor \\
6. Observer (3D+) & Probabilistic $\{0,?\}$ & 4 & temporal\_obs, eterno\_obs \\
\bottomrule
\end{tabular}
\caption{72 semantic primitives across 6 algebraic layers.}
\label{tab:primitives}
\end{table}

Each primitive is assigned a unique prime number, a bit position, a layer, a set of dependencies (parents in the DAG), and a natural-language definition.

\subsection{The Dependency DAG}

Primitives are connected by a DAG with 133+ directed edges encoding the relation ``X requires Y to exist.'' Key structural properties:

\begin{itemize}[nosep, leftmargin=*]
    \item \textbf{Layer monotonicity}: all dependencies of a layer-$n$ primitive lie in layers $\leq n$.
    \item \textbf{3 roots}: vac\'{i}o (bit 0), informaci\'{o}n (bit 1), uno (bit 44)---all layer~1.
    \item \textbf{Transitive closure}: the maximum ancestor set size is 22 (for creador\_obs, layer~6).
\end{itemize}

\subsection{The 14 Dualities}

The 14 dualities emerge in a specific order, each anchored by 2--4 primitives:

\begin{table}[ht]
\centering
\small
\begin{tabular}{clll}
\toprule
\textbf{\#} & \textbf{Duality} & \textbf{Anchors} & \textbf{Layer} \\
\midrule
D1  & Existir / No-existir     & vac\'{i}o, informaci\'{o}n, uno           & 1 \\
D2  & Aqu\'{i} / No-aqu\'{i}  & eje\_profundidad, eje\_vertical            & 2--4 \\
D3  & Antes / Despu\'{e}s     & posici\'{o}n\_temporal, flujo\_temporal    & 3 \\
D4  & Posible / Imposible      & puede, debe, tal\_vez                     & 4 \\
D5  & Identidad / Diferencia   & muchos, tipo\_de, todo, algunos           & 2--4 \\
D6  & Movimiento / Quietud     & mover, hacer, quietud                     & 3 \\
D7  & Orden / Caos             & orden, caos, bien, mal, verdad, mentira   & 3--4 \\
D8  & Uno / Muchos             & uno, muchos                               & 1--2 \\
D9  & Dentro / Fuera           & contenci\'{o}n, uni\'{o}n, separaci\'{o}n & 2 \\
D10 & Parte / Todo             & parte\_de, todo, algunos                  & 2--4 \\
D11 & Vida / Muerte            & vida, muerte, placer, dolor               & 5 \\
D12 & Causa / Efecto           & porque, si\_entonces                      & 3 \\
D13 & Semejante / Diferente    & tipo\_de                                  & 4 \\
D14 & Sujeto / Objeto          & consciente, ausente, temporal\_obs         & 5--6 \\
\bottomrule
\end{tabular}
\caption{The 14 proposed dualities with anchor primitives and layer assignments.}
\label{tab:dualities}
\end{table}

\subsection{Algebraic Layer Chain}

Each layer operates with a distinct algebra. We conjecture that each transition reflects a hierarchy of conceptual complexity, where higher layers require cognitive access to distinctions formalized in lower layers. This is a \emph{semantic} claim, not a claim about the logical independence of formal systems (Kripke defined modal semantics without fuzzy logic; \L{}ukasiewicz defined trivalent logic without modality). The conjecture is partially supported computationally: 12/14 dualities (86\%) are correctly predicted by their algebra assignment.

% ============================================================
\section{Domain Validation}

\subsection{The IDVS Metric}

We define the \textbf{Information Domain Validity Score} for a domain~$d$ as a weighted composite of six sub-metrics:

$$\text{DVS}(d) = \sum_{i=1}^{6} w_i \cdot M_i(d)$$

\noindent with sub-metrics:
\begin{itemize}[nosep, leftmargin=*]
    \item \textbf{M1 (Coverage)} = $1 - \frac{\text{nulls in layers 1--4}}{\text{total in layers 1--4}}$. Penalizes domains with many unclassifiable foundational primitives.
    \item \textbf{M2 (Direct ratio)} = fraction of primitives classified as D. Genuine domains map more primitives directly.
    \item \textbf{M3 (Strong duality)} = fraction of the 12 dual axes where both members are D or A. Measures structural engagement.
    \item \textbf{M4 (DAG consistency)} = $1 - \frac{\text{violations}}{\text{total edges}}$. A violation occurs when a child has higher state than its parent ($D > A > N$).
    \item \textbf{M5 (Anchor consistency)} = agreement between predicted and actual D-A-N on held-out anchor terms.
    \item \textbf{M6 (Layer coherence)} = mean per-layer consistency of D-A-N classifications. Penalizes layers where primitives are incoherently split between engaged and null.
\end{itemize}
\noindent Default weights: $w = (0.25, 0.20, 0.15, 0.15, 0.10, 0.15)$.

\subsection{Classification Protocol}

Each of 72 primitives is classified in each domain as:
\begin{itemize}[nosep, leftmargin=*]
    \item \textbf{D} (Direct): the primitive operates directly in this domain.
    \item \textbf{A} (Analogical): the primitive applies by metaphor or analogy.
    \item \textbf{N} (Null): the primitive does not apply.
\end{itemize}

\subsection{Results}

All 8 positive domains achieve DVS~$\geq 0.719$ under the composite metric. All 3 negative controls fall below the 0.50 threshold:

\begin{table}[ht]
\centering
\small
\begin{tabular}{llccc}
\toprule
\textbf{Domain} & \textbf{Type} & \textbf{DVS} & \textbf{Result} & \textbf{$\sigma$ below mean} \\
\midrule
Psychology  & Positive  & 0.876 & \checkmark & --- \\
Linguistics & Positive  & 0.857 & \checkmark & --- \\
Philosophy  & Positive  & 0.846 & \checkmark & --- \\
Biology     & Positive  & 0.831 & \checkmark & --- \\
Chemistry   & Positive  & 0.803 & \checkmark & --- \\
Music       & Positive  & 0.756 & \checkmark & --- \\
Physics     & Positive  & 0.755 & \checkmark & --- \\
Mathematics & Positive  & 0.719 & \checkmark & --- \\
\midrule
\textbf{Alchemy}    & \textbf{Negative} & \textbf{0.486} & $\times$ & 6.0$\sigma$ \\
\textbf{Astrology}  & \textbf{Negative} & \textbf{0.438} & $\times$ & 6.9$\sigma$ \\
\textbf{Phrenology} & \textbf{Negative} & \textbf{0.322} & $\times$ & 9.1$\sigma$ \\
\bottomrule
\end{tabular}
\caption{DVS results across 8 positive domains and 3 negative controls. Gap between worst positive (Mathematics, 0.719) and best negative (Alchemy, 0.486) is 0.233. All weight configurations maintain discrimination (Section~\ref{sec:sensitivity}).}
\label{tab:idvs}
\end{table}

\subsection{Null Baseline}

To calibrate the threshold, we computed 10,000 random D-A-N assignments preserving the marginal distribution of each domain. The null distribution yields: mean max DVS~=~0.861, $p_{95}$~=~0.931, $p_{99}$~=~0.955. All 8 positive domains exceed $p_{99}$. Note that the composite DVS (v2, 6 sub-metrics) is more informative than the original 2-metric IDVS: the original IDVS assigned perfect 1.000 to all positives, masking genuine variation. The composite DVS reveals that Psychology (0.876) and Linguistics (0.857) are the strongest domains while Mathematics (0.719) is weakest---consistent with the framework's linguistic origins.

\subsection{Sensitivity Analysis}
\label{sec:sensitivity}

The composite DVS is robust to weight variation: all 7 tested weight configurations (default, equal, M1-heavy, M6-heavy, no-M1, no-M6, legacy) maintain perfect discrimination between all 8 positives and all 3 negatives. Leave-one-out analysis shows that dropping any single sub-metric preserves discrimination (minimum gap = 0.147 when dropping M2). No single metric alone suffices---M5 alone fails to discriminate (Alchemy M5 = 0.961 exceeds some positives)---confirming the necessity of the composite approach.

\begin{table}[ht]
\centering
\small
\begin{tabular}{llrl}
\toprule
\textbf{Perturbation} & \textbf{Target} & \textbf{Mean $\Delta$DVS} & \textbf{Large disruptions} \\
\midrule
Remove edge  & DAG topology        & $-0.003$ & 3.4\% \\
Move layer   & Layer assignment    & $0.000$  & 0.0\% \\
Flip D-A-N   & Classification logic & $-0.060$ & 71.4\% \\
\bottomrule
\end{tabular}
\caption{Sensitivity analysis. Robustness gradient: Topology $>$ Layering $>$ Logic.}
\label{tab:sensitivity}
\end{table}

\subsection{Bayesian Model Comparison}

We compared 4 orderings of the 14 dualities using triangularity of the dependency matrix as the test statistic:

\begin{table}[ht]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{Model} & \textbf{Triangularity} & \textbf{BF vs.\ Reversed} \\
\midrule
\textbf{Proposed} & \textbf{0.539} & \textbf{121.4} (decisive) \\
Hegel    & 0.528 & --- \\
Peirce   & 0.484 & --- \\
Reversed & 0.143 & 1.0 \\
\bottomrule
\end{tabular}
\caption{Bayesian model comparison of duality orderings.}
\label{tab:bayesian}
\end{table}

Against 1,000 random orderings ($\mu = 0.338$), the proposed ordering is significant ($p = 0.005$).

% ============================================================
\section{Neural Validation}

\subsection{Architecture}

We augment GPT-2 Medium (355M parameters) with a \textbf{triadic head}: a single linear projection from the 1024-dimensional hidden state to a 72-bit semantic vector. The activation function is iFSQ ($2\sigma(1.6x) - 1$), which distributes activations uniformly across $[-1, +1]$.

\noindent\textbf{Total trainable parameters (v5\_frozen)}: 73,728 (0.02\% of model). The GPT-2 backbone is frozen.

\subsection{Training}

\begin{table}[ht]
\centering
\small
\begin{tabular}{ll}
\toprule
\textbf{Component} & \textbf{Configuration} \\
\midrule
Loss       & $\mathcal{L}_\text{lang} + 0.05 \cdot (\mathcal{L}_\text{tri} + 2\mathcal{L}_\text{sup} + 5\mathcal{L}_\text{sub})$ \\
$\mathcal{L}_\text{tri}$  & diversity + InfoNCE + entropy + alignment \\
$\mathcal{L}_\text{sup}$  & MSE(projection, gold\_target) \\
$\mathcal{L}_\text{sub}$  & $\text{relu}(\text{hypernym}_{01} - \text{hyponym}_{01}).\text{mean}()$ \\
Optimizer  & AdamW ($3 \times 10^{-4}$, $\beta = 0.9/0.95$) \\
Schedule   & Cosine decay, 50\% triadic warmup \\
Data       & WikiText-103 ($\sim$100M tokens) \\
Steps      & 50,000 \\
Hardware   & NVIDIA RTX 4060 Ti 16GB \\
\bottomrule
\end{tabular}
\caption{Training configuration for v5\_frozen.}
\label{tab:training}
\end{table}

\subsection{Training Run Summary}

\begin{table}[ht]
\centering
\small
\begin{tabular}{lcrrcrl}
\toprule
\textbf{Run} & \textbf{Bits} & \textbf{Acc.} & \textbf{Unique} & \textbf{Coh.} & \textbf{PPL} & \textbf{Key Change} \\
\midrule
v1 & 65 & $\sim$90\% & ? & ? & 2,777 & Baseline \\
v2 & 65 & 95.8\% & 3.8\% & ? & 6,735 & Bad targets \\
v3 & 65 & 90.9\% & 89.2\% & 0.653 & 5,388 & Fixed: 65 prims \\
v4 & 72 & 92.5\% & 100\% & 0.876 & 5,608 & +7 primitives \\
\textbf{v5} & \textbf{72} & \textbf{90.8\%}$^\dagger$ & \textbf{100\%} & \textbf{0.818} & \textbf{31.95} & \textbf{Frozen base} \\
\textbf{v6} & \textbf{72} & \textbf{80.9\%} & \textbf{21.4\%} & \textbf{0.643} & \textbf{31.95} & \textbf{2{,}166 concepts} \\
\bottomrule
\end{tabular}
\caption{Training run progression. V5 solves catastrophic forgetting. V6 scales to 2{,}166 concepts. $^\dagger$Majority-class baseline for V5 is 90.3\% (see Table~\ref{tab:v6}).}
\label{tab:runs}
\end{table}

\subsection{Key Finding: Field Density $>$ Individual Targets}

V4 added 7 new primitives to differentiate collapsed dual pairs (Jaccard $> 0.80$). The result was surprising: collapsed pairs resolved even though their individual gold targets did not change. The mechanism is \textbf{field density}---more diverse training examples teach the model what good differentiation looks like.

\subsection{Key Finding: Catastrophic Forgetting Solved}

All runs v1--v4 showed catastrophic language degradation (PPL $> 2{,}700$ vs.\ baseline 31.95). V5 froze the GPT-2 backbone, training only the 73K-parameter triadic head. Result: PPL~=~31.95 (identical to baseline) at a cost of only $-1.7\%$ accuracy.

\subsection{Relational Structure (Regla de Tres)}

The model preserves proportional relationships between dual pairs:

\begin{table}[ht]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{Analogy} & \textbf{Cosine} & \textbf{Bit Acc.} \\
\midrule
good:evil = truth:lie                    & 0.998 & 1.000 \\
creation:destruction = life:death        & 0.998 & 1.000 \\
freedom:control = individual:collective  & 0.995 & 0.986 \\
pleasure:pain = conscious:absent         & 0.996 & 1.000 \\
union:separation = order:chaos           & 0.997 & 1.000 \\
creation:destruction = union:separation  & 0.997 & 1.000 \\
receptive:creator\_obs = truth:lie       & 0.996 & 0.986 \\
move:stillness = mortal\_obs:eternal\_obs & 0.989 & 0.931 \\
\bottomrule
\end{tabular}
\caption{Regla de Tres: relational structure preservation (v5\_frozen). Mean cosine: 0.997 across 8 quadruples.}
\label{tab:r3}
\end{table}

\subsection{Neural Probes}

8 falsifiable questions were tested using reptimeline checkpoint analysis:

\begin{table}[ht]
\centering
\small
\begin{tabular}{clll}
\toprule
\textbf{Probe} & \textbf{Question} & \textbf{Result} & \textbf{$p$-value / detail} \\
\midrule
Q1  & Layer order in representations?           & \textbf{Null} (cascade, not order)  & 0.713 \\
Q2  & Dual anti-correlation?                    & \textbf{5/13 significant}           & $<0.018$ (BH) \\
Q3  & DAG recovery?                             & \textbf{Above random}               & 0.001 \\
Q4  & Triadic dependencies?                     & Null                                & 1.000 \\
Q5  & Phase transition?                         & \textbf{Yes, step 6000}             & 0.0004 \\
Q6  & Unsupervised discovery?                   & Null                                & 1.000 \\
Q7  & Causal verification?                      & Null                                & 1.000 \\
Q13 & Algebraic Markov blanket (FTA-guaranteed) & \textbf{PASS 100\%}                 & 0/500K pairs violate \\
Q13 & MI between coprime interiors              & v8 NEG / v9 MIXED                   & T2 PASS, T4 FAIL \\
Q18 & Quaternionic product axioms               & \textbf{VERIFIED} (T1--T7)          & all axioms hold \\
\bottomrule
\end{tabular}
\caption{Neural probe results. The model learns relational structure (Q2, Q3, Q5); the algebraic structure imposed by unique prime assignment is preserved by construction (Q13 algebraic) and lifts to the quaternionic product (Q18). Statistical recovery of DAG edges via partial correlations (Q13 MI) is only partial in v9 and absent in v8.}
\label{tab:probes}
\end{table}

\noindent\textbf{Interpretation}: The model learns \emph{relational} structure (analogies, some dual axes, DAG edges) but does not recover \emph{ordinal} structure (layer ordering, ternary dependencies). Q13 algebraic is a verification of an FTA-guaranteed property (gcd of coprime interiors $=1$ by construction of the unique prime assignment), confirming implementation correctness rather than emergent structure. Q18 shows that the quaternionic algebra defined over the learned coordinates closes under Hamilton product with the expected non-commutativity. Q13 MI shows that this algebraic structure does not translate to statistical d-separation between bit pairs, consistent with the frozen-base design: the head projects semantics into bit space without restructuring the backbone's internal representations.

\subsection{V6: Extended Corpus (2{,}166 Concepts)}

V5 trained on the 72 core primitives. V6 extends the gold file to 2{,}166 concepts (72 primitives + 2{,}094 derived from a first-principles corpus spanning 14 scientific domains), testing whether the algebraic structure scales beyond the primitive vocabulary.

\begin{table}[ht]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{Metric} & \textbf{v5\_frozen (72)} & \textbf{v6 (2{,}166)} \\
\midrule
Bit accuracy (test)    & 90.8\%$^\dagger$  & 80.9\% \\
Majority baseline      & 90.3\%  & --- \\
Subsumption accuracy   & 94.2\%  & 95.2\% \\
PPL                    & 31.95   & 31.95 \\
Active bits            & 4/72    & 48/72 \\
Dead bits              & 68      & 37 \\
Entropy (bits)         & 0.039   & 0.370 \\
Unique signatures      & 100\%   & 21.4\% \\
Regla de Tres (cosine) & 0.997   & 0.998 \\
Discovered deps        & 174     & 501 \\
Triadic deps           & 918     & 1{,}714 \\
Training time          & 4.3h    & 12.3h \\
\bottomrule
\end{tabular}
\caption{V5 vs V6 comparison. V6 trades individual accuracy for richer structural discovery. $^\dagger$V5's 90.8\% bit accuracy barely exceeds the majority-class baseline (90.3\%) because gold targets are sparse (mean activation rate = 13\%); most of the apparent accuracy comes from correctly predicting inactive bits. V6's 80.9\% is more meaningful given higher bit utilization.}
\label{tab:v6}
\end{table}

\noindent\textbf{Key findings}:

\begin{enumerate}[nosep, leftmargin=*]
    \item \textbf{Backbone intact}: PPL~=~31.95 after 100K steps---freeze-base eliminates catastrophic forgetting even at 2$\times$ training duration.
    \item \textbf{Field density effect}: 30$\times$ more concepts force the model to activate 48 bits (vs.\ 4 in v5), increasing per-bit entropy by 9.5$\times$. The bit space is utilized, not wasted.
    \item \textbf{Semantic clustering}: Extended concepts collapse into 463 unique signatures. Collapses are semantically coherent (e.g., algebra axioms share a single code). The model functions as a \emph{family classifier} rather than an individual discriminator at this scale.
    \item \textbf{Cascade activation}: High-resolution reptimeline analysis (5-step resolution, telescopic X-ray across two architectures) reveals that the phase transition is not instantaneous but a \emph{cascade}: bits activate in waves of 1--8 over $\sim$45 steps (v9, GPT-Neo~125M) or $\sim$145 steps (v8, GPT-2~Medium). In v9, middle-layer primitives (L3--L5: time, moral, body) activate first, while ontological extremes (L1: existence; L6: observer) crystallize last. The theoretical ordering (L1$\to$L6) is not recovered---nor is any fixed layer sequence, as the cascade order differs between architectures. What \emph{is} conserved cross-architecture is the cascade pattern itself: gradual wave propagation, not simultaneous activation.
    \item \textbf{Phase transition}: A sharp transition occurs at the warmup boundary, with entropy jumping, code churn spiking, and bit utilization increasing. At low checkpoint resolution ($\Delta$5K steps) this appears instantaneous (churn~0$\to$1.0); at high resolution ($\Delta$5 steps) churn reaches only $\sim$0.17 and the transition spreads over dozens of steps as a propagating cascade.
\end{enumerate}

% ============================================================
\section{Discussion}

\subsection{What the Framework Claims}

The Emergent Duality framework makes two testable claims:

\begin{enumerate}[nosep, leftmargin=*]
    \item \textbf{Structural claim}: 72 primitives connected by a specific DAG can classify concepts across any knowledge domain using a ternary (D/A/N) code.
    \item \textbf{Ordering claim}: the 14 dualities emerge in a specific sequence dictated by algebraic dependency.
\end{enumerate}

\subsection{Evidence For}

\begin{itemize}[nosep, leftmargin=*]
    \item \textbf{Structural claim}: supported. 8/8 positive domains achieve DVS~$\geq 0.719$, while all 3 negative controls (Astrology 0.438, Alchemy 0.486, Phrenology 0.322) fall below the 0.50 threshold. Discrimination is robust across all tested weight configurations and leave-one-out metric ablations.
    \item \textbf{Ordering claim}: partially supported. Bayesian comparison yields BF~=~121.4 (decisive) against reversed ordering, but only weak evidence against Hegel (BF~=~1.14) and Peirce (BF~=~1.95). The neural model does not recover any fixed layer ordering (Q1, $p$~=~0.71); instead, high-resolution X-ray reveals a cascade where activation order varies by architecture. The cascade pattern itself (gradual wave propagation, scaffold bits, middle layers leading) is conserved across models, but the specific bit-to-layer sequence is not universal.
\end{itemize}

\subsection{Honest Limitations}

\begin{enumerate}[nosep, leftmargin=*]
    \item \textbf{Self-evaluated classifications}: All D-A-N assignments were made by the framework author. Independent human inter-rater data (K2 protocol, human vs.\ human) exists for only 1/9 domains (Mathematics). A complementary probe (Q32) computed Cohen's $\kappa$ between expert classifications and a model-derived second rater (activation-rate thresholding) across 8 of 9 domains. All values fall in the slight-to-poor range ($\kappa \in [-0.082, 0.111]$), with Philosophy highest and Chemistry lowest. This indicates the trained 72-bit code does not recover the expert D-A-N distinctions and \emph{does not substitute} for human inter-rater validation. The permutation tests validate internal consistency; external agreement remains the critical open task.
    \item \textbf{Underpowered rank correlations}: With $n = 7$ (circle) or $n = 14$ (combined), Spearman correlations cannot achieve significance at $\alpha = 0.05$ even for moderate effects.
    \item \textbf{DVS variation across positives}: The composite 6-metric DVS reveals genuine variation among positive domains (0.719--0.876), with Mathematics weakest and Psychology strongest. The original 2-metric IDVS assigned 1.000 to all positives, masking this variation. While the composite metric is more informative, the gap between worst positive (0.719) and best negative (0.486) is only 0.233---a second positive domain near the Mathematics end could narrow this margin.
    \item \textbf{Neural probes}: 5/8 probes returned null results in v8. Cross-architecture replication (v9, GPT-Neo~125M) confirms Q1 and Q2 remain null, while Q5 (phase transition) replicates as positive ($p$~=~0.018). A new probe Q9 (structural conservation) shows topology is conserved cross-model ($>$90\% agreement on active ratio, duals, deps) but S/C/E/A composition diverges ($\chi^2$ $p<$0.001). The null Q1 result is now understood: high-resolution X-ray reveals that emergence follows a cascade pattern, not a fixed layer ordering. The model does not recover L1$\to$L6 because no such universal ordering exists during training---what exists is architecture-dependent wave propagation through the bit space.
    \item \textbf{Scale--algebra tradeoff}: Larger models (355M) show lower subsumption rates (76.9\%) than smaller ones (40M, 98.3\%).
\end{enumerate}

\subsection{Future Directions}

This work opens several lines of investigation:

\begin{enumerate}[nosep, leftmargin=*]
    \item \textbf{Independent rater validation}: The most critical next step. Blind inter-rater studies across all 9 domains would transform self-evaluated D-A-N classifications into externally validated ones. Preliminary data for Mathematics (1 rater) shows agreement; 8 domains remain.
    \item \textbf{Scale--algebra tradeoff}: Understanding why 355M-parameter models show lower subsumption (76.9\%) than 40M models (98.3\%) may reveal fundamental constraints on discrete algebraic learning at scale.
    \item \textbf{Cross-lingual validation}: All primitives are currently named and evaluated in a Spanish-language conceptual frame. Testing whether the same DAG structure holds when primitives are derived independently from other linguistic traditions would strengthen universality claims.
    \item \textbf{Relational prime chains}: Extending the triadic head with compositional prime chain verification for O(1) anti-hallucination---a natural application of the algebraic structure to production AI systems.
    \item \textbf{Expanded corpus training}: V5 was trained on WikiText-103. Training on multilingual or domain-specific corpora could reveal whether the 72-primitive structure generalizes beyond English-language distributional semantics.
    \item \textbf{Additional dualities}: Preliminary analysis of the first-principles corpus (92 files across 14 scientific domains) suggests candidate dualities beyond the 14 validated here. A specific candidate is formally proposed in the framework documentation as \textbf{D15: Collapse/Superposition}---the missing internal operator of layer~6 (probabilistic algebra $\{0,?\}$), with \textit{temporal\_obs} and \textit{eterno\_obs} reassigned as anchors. D15 closes the layer pattern from $1{,}1{,}2{,}2{,}1{,}0$ to the symmetric $1{,}1{,}2{,}2{,}1{,}1$ but is not yet empirically validated. Ongoing training runs are evaluating this and other candidates against the full primitive dataset; results will be incorporated in future revisions.
\end{enumerate}

% ============================================================
\subsection{Conditions for Falsification}
\label{sec:falsification}

We pre-register four conditions under which the Emergent Duality framework should be considered falsified or require fundamental revision:

\begin{enumerate}[nosep, leftmargin=*]
    \item \textbf{Inter-rater failure} (K2): If independent raters achieve Fleiss' $\kappa < 0.40$ on D-A-N classifications across $\geq$4 domains, the framework's core claim---that primitives can be reliably classified into Direct, Analogical, or Null categories---is not reproducible. \emph{Action}: retract universality claims; re-characterize layers as ``proposed'' rather than ``validated''; redesign D-A-N codebook.

    \item \textbf{Negative control passes}: \emph{Resolved.} Both Alchemy (DVS = 0.486) and Phrenology (DVS = 0.322) were evaluated and correctly fall below the 0.50 threshold, joining Astrology (0.438). Discrimination holds across all weight configurations. This criterion is currently satisfied; it would be triggered if a future pseudo-scientific domain achieved DVS~$\geq 0.50$.

    \item \textbf{Cross-linguistic failure}: If the same DAG structure, when independently derived from a typologically distant language (e.g., Mandarin, Swahili), produces $<$40\% overlap with the current 72-primitive inventory, the framework reflects Spanish-language conceptual biases rather than universal structure. \emph{Action}: restrict claims to ``Western philosophical tradition'' rather than ``universal ontology.''

    \item \textbf{Neural ordering recovery}: If a model trained with the full four-component triadic loss on a corpus $\geq$10$\times$ larger than WikiText-103 still fails to recover layer ordering (Q1 probe $p > 0.05$), the proposed emergence sequence is not reflected in distributional semantics at any accessible scale. \emph{Action}: recharacterize the ordering as a philosophical hypothesis rather than an empirically supported claim.
\end{enumerate}

These criteria are pre-registered before the planned inter-rater study (K2) to prevent post-hoc threshold adjustment.

% ============================================================
\section{Related Work}

\begin{itemize}[nosep, leftmargin=*]
    \item \textbf{Wierzbicka (1996)}: Natural Semantic Metalanguage proposes $\sim$65 universal primitives but lacks formal algebraic structure or computational validation.
    \item \textbf{ConceptNet} \citep{speer2017conceptnet}: Large-scale knowledge graph with 34 relation types but no emergence hierarchy.
    \item \textbf{FrameNet} \citep{baker1998framenet}: Frame-based semantics covering $\sim$1,200 frames but domain-specific rather than universal.
    \item \textbf{Category theory approaches} \citep{coecke2010mathematical}: DisCoCat and related formalisms provide compositional semantics but do not address emergence ordering.
\end{itemize}

Our framework differs in proposing (a)~a specific algebraic chain governing emergence, (b)~empirical validation across 9 domains, and (c)~neurosymbolic learnability from natural language.

% ============================================================
\section{Conclusion}

We have presented a candidate ontological framework of 14 proposed dualities organized across 6 algebraic layers, grounded in 72 semantic primitives. The framework achieves perfect IDVS scores across 8 disciplines while correctly rejecting a negative control, with statistical hardening via permutation tests, null baselines, and Bayesian model comparison (decisive against reversed ordering, but only anecdotal against alternative philosophical orderings). A neurosymbolic model (GPT-2 + triadic head) learns the proposed structure at 90.8\% accuracy without degrading language ability, exhibiting near-perfect relational preservation (cosine 0.996). The neural validation is mixed: 3/9 statistical probes reach significance (Q2, Q3, Q5); the two most theoretically important---layer ordering ($p$~=~0.71) and triadic dependencies ($p$~=~1.0)---are null, and the layer order has been reformulated as a cascade rather than an ordering. The algebraic Markov blanket and quaternionic product axioms (Q13 algebraic, Q18) hold by construction over the learned codes. A model-derived second rater (Q32) does not recover expert D-A-N classifications ($\kappa \leq 0.11$ across 8 domains), so the trained model cannot substitute for human inter-rater validation.

The main open question is external validation: independent human evaluators must confirm the D-A-N classifications before universality claims can be considered robust. We release all data, scripts, and model code to enable replication.

\subsection*{Software and Data Availability}

All code and data are available under BUSL-1.1 at:
\begin{itemize}[nosep, leftmargin=*]
    \item Validation scripts, DAG data, and domain analyses: \texttt{triadic-emergent-duality} repository.
    \item Neurosymbolic training pipeline (GPT-2 + triadic head): \texttt{triadic-microgpt} repository.
    \item Representation lifecycle tracking: \texttt{reptimeline} (also available on PyPI).
    \item Triadic projection head (standalone): \texttt{triadic-head} (also available on PyPI).
    \item Pre-trained weights (V5\_frozen GPT-2 Medium, production milestone): \href{https://huggingface.co/arturoornelasb/triadic-gpt2-medium-v5}{huggingface.co/arturoornelasb/triadic-gpt2-medium-v5}
    \item Pre-trained weights (V6 GPT-2 Medium, extended corpus 2{,}166 concepts): \href{https://huggingface.co/arturoornelasb/triadic-gpt2-medium-v6}{huggingface.co/arturoornelasb/triadic-gpt2-medium-v6}
    \item Pre-trained weights (V8 GPT-2 Medium, deep head + X-ray base): \href{https://huggingface.co/arturoornelasb/triadic-gpt2-medium-v8}{huggingface.co/arturoornelasb/triadic-gpt2-medium-v8}
    \item Pre-trained weights (V9 GPT-Neo 125M, cross-architecture replication): \href{https://huggingface.co/arturoornelasb/triadic-gptneo-125m-v9}{huggingface.co/arturoornelasb/triadic-gptneo-125m-v9}
\end{itemize}
The repository also includes a structured compilation of first principles across 14 scientific domains (92 files organized hierarchically from mathematics to interdisciplinary fields), used by the algebraic gap analysis to search for candidate operators in uncovered inter-layer connections.

\subsection*{Companion Papers}

This work is part of a multi-paper program on triadic neurosymbolic representations. The present paper's claims are evaluated independently; companion results provide additional context but are not required for the conclusions drawn here. Each companion component has its own publication and repository:

\begin{itemize}[nosep, leftmargin=*]
    \item \textbf{Triadic Neurosymbolic Engine}: Ornelas Brand, J.~A. (2026). \emph{Prime Factorization as a Neurosymbolic Bridge: Projecting Continuous Embeddings into Discrete Algebraic Space for Deterministic Verification.} Paper: \href{https://doi.org/10.5281/zenodo.19205805}{10.5281/zenodo.19205805}. Repository: \href{https://doi.org/10.5281/zenodo.18748671}{10.5281/zenodo.18748671}.
    \item \textbf{reptimeline}: Ornelas Brand, J.~A. (2026). \emph{Tracking Discrete Representation Evolution During Neural Network Training.} Paper: \href{https://doi.org/10.5281/zenodo.19208672}{10.5281/zenodo.19208672}. Repository: \href{https://doi.org/10.5281/zenodo.19208628}{10.5281/zenodo.19208628}.
    \item \textbf{triadic-microgpt}: Ornelas Brand, J.~A. (2026). \emph{End-to-End Prime Factorization in a Generative Language Model: Learned Algebraic Encoding from Joint Training.} Paper: \href{https://doi.org/10.5281/zenodo.19206545}{10.5281/zenodo.19206545}. Repository: \href{https://doi.org/10.5281/zenodo.19207845}{10.5281/zenodo.19207845}.
    \item \textbf{Duality Synthesis in Quaternionic Logic}: Ornelas Brand, J.~A. (2026). \emph{How Opposites Generate Truth.} Paper: \href{https://doi.org/10.5281/zenodo.19561634}{10.5281/zenodo.19561634}. Defines synthesis $a \oplus \bar{a} = \mathrm{lcm}(\Phi(a),\Phi(\bar{a}))$ on the prime algebra and verifies the quaternionic product axioms exercised in Q18 above.
    \item \textbf{Pre-Logical States and the Birth of Information}: Ornelas Brand, J.~A. (2026). \emph{How Reasoning Operates Before Truth Exists.} Paper: \href{https://doi.org/10.5281/zenodo.19561722}{10.5281/zenodo.19561722}. Formalizes pre-logical states in the quaternionic framework, connecting collapse to Friston's free energy minimization.
    \item \textbf{The Adjacent Possible Has a Lattice} (in preparation): Ornelas Brand, J.~A. (2026). Formalizes Kauffman's adjacent possible via prime product spaces. Supplementary data in \texttt{data/adjacent\_possible\_paper/}.
\end{itemize}

% ============================================================
\appendix
\section{Notation}

\begin{table}[ht]
\centering
\small
\begin{tabular}{ll}
\toprule
\textbf{Symbol} & \textbf{Meaning} \\
\midrule
D, A, N & Direct, Analogical, Null classification states \\
IDVS    & Information Domain Validity Score \\
BF      & Bayes Factor \\
PPL     & Perplexity \\
iFSQ    & Improved Finite Scalar Quantization \\
BH-FDR  & Benjamini--Hochberg False Discovery Rate \\
SHD     & Structural Hamming Distance \\
DAG     & Directed Acyclic Graph \\
\bottomrule
\end{tabular}
\caption{Notation reference.}
\label{tab:notation}
\end{table}

% ============================================================
\bibliographystyle{plainnat}

\begin{thebibliography}{7}

\bibitem[Wierzbicka(1996)]{wierzbicka1996}
A.~Wierzbicka.
\newblock \emph{Semantics: Primes and Universals}.
\newblock Oxford University Press, 1996.

\bibitem[Speer et~al.(2017)]{speer2017conceptnet}
R.~Speer, J.~Chin, and C.~Havasi.
\newblock ConceptNet 5.5: An open multilingual graph of general knowledge.
\newblock In \emph{AAAI}, 2017.

\bibitem[Baker et~al.(1998)]{baker1998framenet}
C.~F. Baker, C.~J. Fillmore, and J.~B. Lowe.
\newblock The Berkeley FrameNet project.
\newblock In \emph{COLING-ACL}, 1998.

\bibitem[Coecke et~al.(2010)]{coecke2010mathematical}
B.~Coecke, M.~Sadrzadeh, and S.~Clark.
\newblock Mathematical foundations for a compositional distributional model of meaning.
\newblock \emph{Linguistic Analysis}, 2010.

\bibitem[Radford et~al.(2019)]{radford2019language}
A.~Radford, J.~Wu, R.~Child, et~al.
\newblock Language models are unsupervised multitask learners.
\newblock \emph{OpenAI Technical Report}, 2019.

\bibitem[Benjamini and Hochberg(1995)]{benjamini1995controlling}
Y.~Benjamini and Y.~Hochberg.
\newblock Controlling the false discovery rate: A practical and powerful approach to multiple testing.
\newblock \emph{Journal of the Royal Statistical Society: Series B}, 57(1):289--300, 1995.

\bibitem[Mentzer et~al.(2025)]{mentzer2025fsq}
F.~Mentzer et~al.
\newblock Finite scalar quantization: {VQ-VAE} made simple.
\newblock In \emph{ICLR}, 2024.

\end{thebibliography}

\end{document}
